Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-04 application.tex
edits in progress to 6 plus some suggestions to 7, not done yet...

\section{Application to protoplanetary disks}\label{application} 
As an application of our linear models, we estimate where in a
PPD do the thermodynamic conditions allow the VSI to
operate. For a realistic disk one might consider an energy equation
with radiative cooling, 
\begin{align}\label{real_energy}
  % \rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
  \frac{\p P}{\p t} + \bm{v}\cdot\nabla P +\gamma P \nabla\cdot\bm{v} = - \frac{P}{\rho T
    C_v}\left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}
where $C_v$ is the heat capacity at constant
volume, 
$\bm{F}$ is the radiative flux and $Q_+$ is a heating
term to permit an equilibrium solution.

@@

\subsection{Radiative diffusion}
For perturbations with lengthscales $l\gg
l_\mathrm{rad}$, we assume temperature fluctuations are smoothed out
by radiative diffusion and model this in linear theory by writing the
linearized RHS of Eq. \ref{real_energy} as

\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T C_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right),  
\end{align}
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly.

@@

\subsection{Newtonian cooling}\label{newton_cool}
For perturbations with small lengthscales, $l\ll l_\mathrm{rad}$, 
the disk is optically thin. In this case, we assume 
Newtonian cooling can be applied, which is the thermal
relaxation model as adopted in our linear models. That is, temperature
fluctuations $\delta T$ decay on a timescale $t_\mathrm{cool}$,
independent of $k_x$. For simplicity, we take  
\begin{align}
  t_\mathrm{cool} = \frac{l_\mathrm{rad}^2}{3\eta}. 
\end{align}
We note that $t_\mathrm{cool}$ does not explicitly depend on $\rho$.

>>>>

\section{Application to protoplanetary disks}\label{application} 
\emph{I suggest the following edits
\begin{itemize}
\item Instead of starting over again with the full energy equation focus on giving a phyiscal values for $\Lambda$ and $t_c$ already introduced.
\item Don't describe what we don't do, especially at the outset.  Say what's done, and then if essential list caveats.
\end{itemize}

As an application of our linear models, we estimate where in a
PPD do the thermodynamic conditions allow the VSI to
operate. For a realistic disk one might consider an energy equation
with radiative cooling, 
\begin{align}\label{real_energy}
  % \rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
  \frac{\p P}{\p t} + \bm{v}\cdot\nabla P +\gamma P \nabla\cdot\bm{v} = - \frac{P}{\rho T
    C_v}\left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}
where $C_v$ is the heat capacity at constant
volume, 
$\bm{F}$ is the radiative flux and $Q_+$ is a heating
term to permit an equilibrium solution.

@@

\subsection{Radiative diffusion}
\emph{the following sentence is too long and has a dangling "this".}

For perturbations with lengthscales $l\gg
l_\mathrm{rad}$, we assume temperature fluctuations are smoothed out
by radiative diffusion and model this in linear theory by writing the
linearized RHS of Eq. \ref{real_energy} as

\emph{I would have this be an equation for something, like $\Gamma$ or $t_c$.}
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T C_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right),  
\end{align}
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly.

\emph{To get to the point more directly and intuitively, I would start below with "We include only the radial derivatives in Eq.\ blah, neglecting vertical derivatives.  This approximation is valid for the vertically elongated modes we consider."  I think we've already discussed vertical elongation enough that this can be brief.  Then maybe say that keeping the vertical derivates would be bad, if you really want to.  "With this simplification (give revised equation)}

@@

\subsection{Newtonian cooling}\label{newton_cool}
For perturbations with small lengthscales, $l\ll l_\mathrm{rad}$, 
the disk is optically thin.
\emph{the disk isn't (necessarily) opt. thin, the pert. is}
 In this case, we assume 
Newtonian cooling can be applied, which is the thermal
relaxation model as adopted in our linear models. That is, temperature
fluctuations $\delta T$ decay on a timescale $t_\mathrm{cool}$,
independent of $k_x$. For simplicity, we take  
\begin{align}
  t_\mathrm{cool} = \frac{l_\mathrm{rad}^2}{3\eta}. 
\end{align}
We note that $t_\mathrm{cool}$ does not explicitly depend on $\rho$.

### 2015-05-04 numerical.tex
edits in progress to 6 plus some suggestions to 7, not done yet...

\subsection{Rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. Since this cooling is much faster than the predicted
$\beta_\mathrm{crit}$, we expect and that this case should give 
similar results to the well studied case of isothermal perturbations
 \citepalias[e.g.][]{nelson13,mcnally14,barker15}.  Indeed we found that smaller
 $\beta$ values gave similar results. This case allows us to explore and test our
finite cooling time model in a familiar context.

@@

\subsubsection{Case study: $\khat=10$}
In Fig. \ref{compare_modes_iso_kx10} we plot eigenvalues -- growth rates and
wave frequencies -- of unstable 
modes for $\khat=10$.  We compare our numerical results to the analytic dispersion relations for 
$\beta = 0$ and $10^{-3}$, from Eqs.\ \ref{simple_growth} and \ref{relax_disp}, respectively. Each symbol 
corresponds to a different order mode, i.e.\ a different number of vertical node crossings.  Lower order modes
have lower (in magnitude) frequencies, with the lowest order `fundamental' modes further to the left in Fig. \ref{compare_modes_iso_kx10}.

@@

Rather, the decline in growth rate for high order models is due to the vertical box size, as seen by comparing the $\zmax/H = 3, 5$ and 7 cases in Fig. \ref{compare_modes_iso_kx10}.  Larger boxes give better agreement with the analytic theory, which are infinite in vertical extent.  Larger boxes include regions of larger vertical shear (see Eqs. \ref{vertical_shear_thin} and \ref{vertical_shear_ex}), which higher order modes can tap to give larger growth rates.   
The need for vertically extended domains to capture the largest VSI growth rates is problematic 
For the $\zmax = 7 H$ case, we begin to see another branch of modes, the so-called ``surface modes" described below.

@@

The modes shown in Fig. \ref{compare_modes_iso_kx10} are `body modes'
and are inertial waves destabilized by vertical shear
\citepalias{barker15}. (The exception being the fastest growing
mode for $\zmax=7H$ which is a surface mode --- see the discussion
below.)  The `fundamental corrugation mode', marked by `f',
corresponds to the eigenvalue with the smallest $|\upsilon|$. The
fundamental mode is robust  
against vertical boundary conditions as it appears at the same
position in Fig. \ref{compare_modes_iso_kx10} for all cases. Higher  
order modes, corresponding to increasing $|\omega|$, increasingly
deviates from the analytic line due to the imposed boundaries, 
unless $\zmax$ is increased.

For later comparison, we plot in Fig. \ref{lowfreq_eigenfunc} the 
eigenfunctions $W$ and $\delta   v_z$ for the fundamental mode. We also show a 
meridional visualization in Fig. \ref{lowfreq_eigenfunc_2d}. The $x$ axis is 
stretched for clarity; radial velocities are in fact typically much 
smaller than vertical velocities. Most of the meridional kinetic energy is
contained within $\sim 2H$ of the midplane due to the density 
stratification.

@@

\subsection{Effect of thermal relaxation}\label{therm_relax_eff}
We now consider the effect of finite thermal relaxation by increasing 
$\beta$. % Note that for the fiducial disk model the critical thermal 
% timescale $\beta_\mathrm{crit} = 0.125$.
Fig. \ref{compare_modes_cool_kx10} show unstable modes with $\khat=10$ 
and $\beta\in[0.01,0.1]$. In addition to our fiducial setup with 
$\zmax=5H$, we also plot results for $\zmax=3H$, $\zmax=7H$ and 
analytic results computed from Eq. \ref{relax_disp}.  %analytic
% results for low order modes

We find the number of unstable modes decrease with increasing
$\beta$. Consider the fiducial setup with $\zmax=5H$ 
(middle panel). The fundamental mode ($M=0$) is not the most unstable for   
$\beta=0.01$ as there area number of modes with $M>0$ that have larger
growth rates. As noted above, these higher order modes are affected by
$\zmax$, e.g. there is a significant mismatch between numerical and 
analytical results for $M\geq1$ when $\zmax=3H$ (upper panel).

@@

Nevertheless, as before we observe more effective stabilization for
higher order modes, including the absence of surface modes, for
example as $\beta$ is increased from $0.01$ to $0.03$ (for $\zmax\geq
5H$). If we discard modes associated with boundary conditions, then
the fundamental mode would again become the dominant mode as $\beta$ is
increased.

@@

\subsection{Vertically non-isothermal disk model} 
%We briefly consider vertically non-isothermal disks in which there is
%a disk surface at finite height. 
\citetalias{barker15} have argued that vertically isothermal disks are not
ideal for studing the VSI because $\zmax$ is a free 
parameter in numerical calcluations. On the other hand, vertically 
non-isothermal disks have a disk surface at finite height so there is
a natural choice for $\zmax$. This means that higher order body modes and
surface modes are better defined in a vertically non-isothermal disk.

@@

Nevertheless, here we briefly consider a vertically non-isothermal
disk model with $(p,q, h)=(0,-1,0.05)$ and $\Gamma=1.0837$. This
value of $\Gamma$ is chosen so that the standard vertical domain size $\zmax =
5H$ effectively corresponds to the disk surface ($\zmax\simeq
H_s$). The disk temperature decreases rapidly with height such that 
$c_s^2(\zmax)\simeq 10^{-4}c_s^2(0)$.

>>>>

\subsection{Rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$.  Since $\beta \ll \beta_\mathrm{crit}$, this case
gives similar results to the well studied case of isothermal perturbations
 \citepalias[e.g.][]{nelson13,mcnally14,barker15}.\footnote{To be sure, we confirmed that smaller
 $\beta$ values gave similar results.} This case allows us to explore and test our
finite cooling time model in a familiar context.

@@

\subsubsection{Case study: $\khat=10$}
In Fig. \ref{compare_modes_iso_kx10} we plot eigenvalues -- growth rates and
wave frequencies -- of unstable 
modes for $\khat=10$.  We compare our numerical results to the analytic dispersion relations for 
$\beta = 0$ and $10^{-3}$, from Eqs.\ \ref{simple_growth} and \ref{relax_disp}, respectively. Each symbol 
corresponds to a different order mode, i.e.\ a different number of vertical node crossings.  Lower order modes
have lower (in magnitude) frequencies for a given radial wavenumber, as expected for inertial-gravity waves (see \S\ref{stable_novshear}).

@@

Rather, the decline in growth rate for high order models is due to the vertical box size, as seen by comparing the $\zmax/H = 3, 5$ and 7 cases in Fig. \ref{compare_modes_iso_kx10}.  Larger boxes give better agreement with the analytic theory, which has infinite vertical extent.  Larger boxes include regions of larger vertical shear (see Eqs. \ref{vshear_thin} and \ref{vertical_shear_ex}), which higher order modes can tap to give larger growth rates.   For the $\zmax = 7 H$ case, we begin to see another branch of modes with the highest growth rates at $\omega \sim -6 h\OmK$. This branch contains the  ``surface modes" described below.

The need for vertically extended domains to capture the largest VSI growth rates is problematic, especially for hydrodynamic simulations.  This  complication is mitigated by at least two factors.  First, we show below that with slower cooling, the growth of higher order modes is preferentially damped, consistent with the analytic analysis in \S\ref{iso_vsi_beta_crit}.  Second, the fastest growing modes may not dominate transport when they operate in very low density  surface layers, i.e. at many $H$.   Quantifying which modes will contribute most to non-linear transport is an important issue, but beyond the scope of this work.

@@

%The modes shown in Fig. \ref{compare_modes_iso_kx10} are `body modes'
%and are inertial waves destabilized by vertical shear
%\citepalias{barker15}. (The exception being the fastest growing
%mode for $\zmax=7H$ which is a surface mode --- see the discussion
%below.)  The `fundamental corrugation mode', marked by `f',
%corresponds to the eigenvalue with the smallest $|\upsilon|$. The
%fundamental mode is robust  
%against vertical boundary conditions as it appears at the same
%position in Fig. \ref{compare_modes_iso_kx10} for all cases. Higher  
%order modes, corresponding to increasing $|\omega|$, increasingly
%deviates from the analytic line due to the imposed boundaries, 
%unless $\zmax$ is increased.

Fig. \ref{lowfreq_eigenfunc} shows 
eigenfunctions $W = \delta P/\rho$ and $\delta   v_z$ for the lowest order fundamental mode.  
Fig. \ref{lowfreq_eigenfunc_2d} maps the pressure perturbation and meridional flow,
scale by $\sqrt{\rho}$ to reflect the contribution to kinetic energy. Due to the stretched $x$ axis, 
radial velocities are in fact typically much 
smaller than vertical velocities, as expected for a vertically elongated, incompressible mode. 
Most of the kinetic energy is contained within $\sim 2H$ of the midplane due to the density 
stratification.

@@

\subsection{Effect of thermal relaxation}\label{therm_relax_eff}
We now consider the effect of finite thermal relaxation by increasing 
$\beta$. % Note that for the fiducial disk model the critical thermal 
% timescale $\beta_\mathrm{crit} = 0.125$.
Fig. \ref{compare_modes_cool_kx10} show unstable modes with $\khat=10$ 
and $\beta$ from $0.01$ to  $0.1$. In addition to our fiducial setup with 
$\zmax=5H$, we also plot results for $\zmax=3H$, $\zmax=7H$ and 
analytic results computed from Eq. \ref{relax_disp}.  %analytic
% results for low order modes

We find the number of unstable modes decrease with increasing
$\beta$. Consider the fiducial setup with $\zmax=5H$ 
(middle panel). The fundamental mode ($M=0$) is not the most unstable for   
$\beta=0.01$ as higher order modes have larger
growth rates. As noted above, these higher order modes are affected by
$\zmax$, e.g. there is a significant mismatch between numerical and 
analytical results for $M\geq1$ when $\zmax=3H$ (upper panel).

@@

Nevertheless, as before we observe more effective stabilization for
higher order modes, including the absence of surface modes, for
example as $\beta$ is increased from $0.01$ to $0.03$ (for $\zmax\geq
5H$). 
%\emph{not a clear and possible not valid argument.  All modes are associated with
%a BC and the disappearance of a mode for smaller zmax could be the artificial effect.}
%If we discard modes associated with boundary conditions, then
%the fundamental mode would again become the dominant mode as $\beta$ is
%increased.

@@

\subsection{Vertically non-isothermal disk model} 
%We briefly consider vertically non-isothermal disks in which there is
%a disk surface at finite height. 
\emph{fix this rambling discussion..}
\citetalias{barker15} have argued that vertically isothermal disks are not
ideal for studing the VSI because $\zmax$ is a free 
parameter in numerical calcluations. On the other hand, vertically 
non-isothermal disks have a disk surface at finite height so there is
a natural choice for $\zmax$. This means that higher order body modes and
surface modes are better defined in a vertically non-isothermal disk.

@@

Nevertheless, here we briefly consider a vertically non-isothermal
disk model with $(p,q, h)=(0,-1,0.05)$ and $\Gamma=1.0837$. This
value of $\Gamma$ is chosen so that the standard vertical domain size $\zmax =
5H$ effectively corresponds to the disk surface ($\zmax\simeq
H_s$). The disk temperature decreases rapidly with height such that 
$c_s^2(\zmax)\simeq 10^{-4}c_s^2(0)$.    \emph{note that this would imply midplane
temperatures in excess of $10^6$ K so it's completely un-astrophyscial.}
